%% CS505 Final Project Proposal
%% Single-column proposal format (ACL format reserved for midway/final reports)
%% Compile: pdflatex → bibtex → pdflatex → pdflatex

\documentclass[11pt,letterpaper]{article}

%% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage[top=0.9in, bottom=0.9in, left=1in, right=1in]{geometry}
\usepackage{natbib}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{parskip}

\makeatletter
\renewcommand{\section}{\@startsection{section}{1}{\z@}%
  {5pt \@plus 1pt \@minus 1pt}{1.5pt}%
  {\normalfont\normalsize\bfseries}}
\renewcommand{\subsection}{\@startsection{subsection}{2}{\z@}%
  {3pt \@plus 1pt}{1pt}%
  {\normalfont\normalsize\itshape}}
\makeatother
\setlength{\parskip}{2.5pt}
\setlength{\parindent}{0pt}
\bibsep=2pt

%% ── Title ─────────────────────────────────────────────────────────────────────
\title{\vspace{-1.5em}%
  \textbf{Reducing Hallucinations via Citation-Grounded QA
  with Locally-Hosted Small Language Models}\\[3pt]
  {\normalsize\normalfont
   CS505 Natural Language Processing \enspace|\enspace
   Final Project Proposal \enspace|\enspace Boston University}\\[1pt]
  {\normalsize\normalfont [We will add names here :)]}%
}
\date{\vspace{-1.5em}}

\begin{document}
\maketitle
\thispagestyle{empty}

%% ── Preamble ─────────────────────────────────────────────────────────────────
\noindent\textbf{Project type:}
Empirical NLP project focused on retrieval-augmented generation (RAG) for
hallucination reduction.

\noindent\textbf{Research question:}
\emph{Within a fully local, privacy preserving RAG system, how much of a
sub-7B SLM's hallucination rate is attributable to retrieval quality versus
generation capacity and can systematic retrieval improvements make small
local models practically reliable for citation-grounded QA over private corpora?}

%% ── 1. Problem Statement ─────────────────────────────────────────────────────
\section{Problem Statement \& Motivation}

Hallucination generating claims unsupported by source documents is the
primary failure mode of LLMs in high stakes settings.
Cloud tools like Google's NotebookLM answer questions over private documents
with inline citations, but require transmitting the entire corpus to an
external server, which may be incompatible with HIPAA/GDPR requirements
or internal data-governance policies.
\textbf{Cloud upload equals data-exfiltration risk.}
Sub-7B SLMs (e.g., Phi-3-mini-3.8B \citep{microsoft2024phi3}) can run on a
single consumer GPU, but it is unclear whether their hallucination rate
stems primarily from \emph{retrieval failures} or \emph{generation failures}.
Disentangling the two is the key to principled, cost-effective improvement.

\textbf{Task definition.}
Given a corpus $\mathcal{C}=\{d_1,\ldots,d_n\}$ and query $q$, the system
produces answer $a$ plus, for every sentence $s_i\in a$, a citation
$(p_i, s_i)$ where passage $p_i\in\mathcal{C}$ textually entails $s_i$.
Any sentence without a valid citation is counted as a hallucination.

%% ── 2. Proposed System ───────────────────────────────────────────────────────
\section{Proposed System}

All components run locally; no external API calls are required.

\textbf{Parsing.}  PyMuPDF; preserves section headings for structure-aware
chunking.

\textbf{Chunking (ablated).}  Three strategies: fixed-size (256 tok, 32
overlap), sliding-window (512 tok, 128 overlap), section-aware
(boundary-aligned, capped at 512 tokens to prevent oversized chunks).

\textbf{Retrieval (ablated).}  Dense-only (FAISS + BGE-small-en-v1.5);
BM25 (rank-bm25); hybrid reciprocal-rank fusion (RRF); hybrid +
cross-encoder reranking (ms-marco-MiniLM-L-6-v2), top-5 passages.

\textbf{Generation.}  Two prompt variants: standard generation (no citation
instruction) vs.\ \emph{citation-forcing} (the model must bracket every claim
with an inline citation and assert only what the retrieved passages support).
Served via Ollama; $\leq$16 GB VRAM.
If Phi-3-mini fails to follow the bracket format consistently, we will fall back
to constrained decoding (Guidance or Outlines) to enforce the schema at the token level.

%% ── 3. Experimental Design ───────────────────────────────────────────────────
\section{Experimental Design}

\subsection{Conditions}

Each condition changes exactly one component (Table~\ref{tab:ablation}).
Conditions A to E are the committed core; F and G (both local) are stretch goals
testing whether retrieval improvements generalize across model families.

\begin{table}[h]
\centering
\small
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{@{}clllcc@{}}
\toprule
\textbf{Cond.} & \textbf{Chunk} & \textbf{Retrieval} &
\textbf{Rerank} & \textbf{Cite-force} & \textbf{Generator} \\
\midrule
A (base) & Fixed   & Dense  & $\times$   & $\times$   & Phi-3-mini \\
B        & Fixed   & Hybrid & $\times$   & $\times$   & Phi-3-mini \\
C        & Section & Hybrid & $\times$   & $\times$   & Phi-3-mini \\
D        & Section & Hybrid & \checkmark & $\times$   & Phi-3-mini \\
E (best) & Section & Hybrid & \checkmark & \checkmark & Phi-3-mini \\
\midrule
F$^*$    & Section & Hybrid & \checkmark & \checkmark & Llama-3-8B \\
G$^*$    & Section & Hybrid & \checkmark & \checkmark & Mistral-7B \\
\bottomrule
\end{tabular}
\caption{Ablation conditions. $^*$Stretch goals (time permitting).}
\label{tab:ablation}
\end{table}

\subsection{Dataset \& Baselines}

\textbf{Dataset:} QASPER \citep{dasigi2021qasper}—1,585 NLP papers,
5,049 Q\&A pairs with gold evidence annotations supporting both
retrieval and answer evaluation (official token F1).
Evidence annotations are treated as approximate grounding signals;
NLI and token overlap together compensate for annotation incompleteness.
We use QASPER as a controlled benchmark for citation-grounded QA; our
target deployment is private corpora where gold-annotated evaluation
of this kind is not feasible.

\textbf{Baselines:}
(1)~\textbf{Parametric-only LM}: Phi-3-mini with no retrieval, establishing
the hallucination floor.
(2)~\textbf{BM25-only} sparse keyword retrieval, the simplest retrieval
baseline.
(3)~\textbf{Random Retrieval}: five passages sampled uniformly at random,
confirming that RRF gains reflect retrieval \emph{quality} rather than
the mere presence of context.

\subsection{Metrics \& Hypothesis}

We treat citation precision as the primary measure of faithfulness;
all other metrics serve as supporting signals or retrieval diagnostics.

\textbf{Primary:} Citation precision via NLI entailment (TRUE cross-encoder)
and hallucination rate; logs \% sentences with valid citations and
\% malformed citation-format outputs per condition.
\textbf{Secondary:} Token F1 (QASPER official), BERTScore.
\textbf{Faithfulness signal:} RAGAS \citep{es2023ragas} with Llama-3-8B as
local judge (Ollama)—supporting signal only, to avoid LLM-judges-LLM circularity.
\textbf{Retrieval diagnostics:} Recall@5, MRR@10, evidence hit rate.
\textbf{Validation:} Human evaluation on 20--30 sampled QA pairs as tie-breaker.

We will additionally plot hallucination rate against Recall@5 across conditions
and stratify results by whether gold evidence was retrieved, directly attributing
failures to retrieval versus generation.

\textbf{Hypothesis:} Retrieval improvements (A$\to$D) will account for the
majority of hallucination reduction, with citation-forcing (D$\to$E) providing
a smaller incremental gain; the retrieval--hallucination correlation will show
a marked increase in faithfulness once Recall@5 crosses a threshold value.

%% ── 4. Related Work ──────────────────────────────────────────────────────────
\section{Related Work}

\citet{lewis2020rag} introduced RAG and remains the canonical reference for
retrieval-augmented generation; \citet{asai2024openscholar} is the closest
prior system, applying RAG to scientific literature synthesis.
Our contribution—a controlled ablation on a locally-hosted SLM that
disentangles retrieval failures from generation failures—is absent from
both works and from the broader attribution evaluation literature.

%% ── 5. Feasibility & Timeline ────────────────────────────────────────────────
\section{Feasibility \& Timeline}

\textbf{Scope.}
The committed core (A to E on QASPER with Phi-3-mini) builds the retrieval
pipeline incrementally—one component per week—and is realistic on a single
consumer GPU.
Stretch goals (F: Llama-3-8B, G: Mistral-7B) will be attempted in
week 5 and deferred to the final report if time is short.

\begin{table}[h]
\centering
\small
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.05}
\begin{tabular}{@{}cp{9cm}@{}}
\toprule
\textbf{Wk} & \textbf{Milestone} \\
\midrule
1--2 & Parsing, chunking strategies, BM25 + dense baseline (cond.\ A) \\
3    & Hybrid RRF + reranking; conditions B to D; QASPER eval pipeline \\
4    & Citation-forcing (E); RAGAS + NLI evaluation; error analysis \\
5    & Midway report; begin stretch goals (F, G) if on schedule \\
6    & Final ablation, writeup (ACL format) \\
\bottomrule
\end{tabular}
\caption{Six-week schedule. Core scope: weeks 1-4 (conditions A to E).}
\end{table}

\noindent\textbf{Stack:} PyMuPDF, sentence-transformers, FAISS, rank-bm25,
Ollama, RAGAS, HuggingFace Transformers; no paid API for core scope.

%% ── References ───────────────────────────────────────────────────────────────
{\footnotesize
\bibliographystyle{abbrvnat}
\bibliography{references}
}

\end{document}
